---
title: 'Reproducible Research: Peer Assessment 1'
author: "Jeremy Voisey"
date: "19 March 2017"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
It is now possible to collect a large amount of data about personal movement
using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone
Up. These type of devices are part of the "quantified self" movement - a group
of enthusiasts who take measurements about themselves regularly to improve
their health, to find patterns in their behavior, or because they are tech geeks.
But these data remain under-utilized both because the raw data are hard to
obtain and there is a lack of statistical methods and software for processing
and interpreting the data.

This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps
taken in 5 minute intervals each day.

## Loading and preprocessing the data
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded
as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a
total of 17,568 observations in this dataset.

The file was read into a data frame, activity

```{r load}
filename <- "activity/activity.csv"
activity <- read.csv(filename, stringsAsFactors = FALSE)
```


## What is mean total number of steps taken per day?

```{r dailysteps}
suppressMessages(library(dplyr))

dailysteps <- activity %>%
    group_by(date) %>%
    summarise(dailysteps = sum(steps, na.rm = TRUE))
```

### The distribution of total daily steps

```{r histogram}
suppressMessages(library(ggplot2))
ggplot(dailysteps, aes(x = dailysteps)) +
    geom_histogram(binwidth = 1000) +
    labs(x = "Daily Steps", title = "Distribution of Total Daily Steps")

```

### The mean and median of total daily steps

```{r meandailysteps}
mean_dailysteps <- round(mean(dailysteps$dailysteps),0)
mean_dailysteps
median_dailysteps <- median(dailysteps$dailysteps)
median_dailysteps
```

The mean number of daily steps was `r mean_dailysteps`,
the median was `r median_dailysteps`

## What is the average daily activity pattern?
```{r dailyactivity}
dailyactivity <- activity %>%
    group_by(interval) %>%
    summarise(averagesteps = mean(steps, na.rm = TRUE))
```

### Average number of steps taken vs. 5-Minute interval
```{r lineplot}
library(ggplot2)
ggplot(dailyactivity, aes(x = interval, y = averagesteps)) +
    geom_line() +
    labs(x = "5 Minute Interval", y = "average number of steps", title = "Average number of steps taken vs. 5-Minute interval")

```

#### Maximum number of steps
```{r max steps}
maxinterval <- dailyactivity[which.max(dailyactivity$averagesteps), "interval"]
maxinterval
```

The interval containing the maximum number of steps on average was `r maxinterval`

## Imputing missing values
```{r missing}
missingsteps <- sum(is.na(activity$steps))
missingsteps
rowcount <- nrow(activity)
rowcount
```

There are `r missingsteps` rows, missing values out of `r rowcount`

Missing values were filled in by assuming that the number of steps in that
interval was equal to the average for that interval across all days

```{r impute}
# Make copy of activity data.frame
filled_activity <- activity
# Identify rows with missing values
whichna <- which(is.na(filled_activity$steps))
# Loop over missing values and replace them with average for corresponding interval
for (rown in whichna) {
    intervalna <- filled_activity$interval[rown]
    filled_activity$steps[rown] <- with(dailyactivity,averagesteps[which(interval == intervalna)]) 
}
```

### What is mean total number of steps taken per day for Imputed data set?
```{r filled_dailysteps}
filled_dailysteps <- filled_activity %>%
    group_by(date) %>%
    summarise(dailysteps = sum(steps, na.rm = TRUE))
```

### The distribution of total daily steps using Imputed data set

Imputing missing values has the noticable effect of decreasing the the number
of days with a count of zero (missing values are effectively treated as 0) and
Increasing the number of days with 11000 steps.

```{r filled_histogram}
combined_dailysteps <- rbind(mutate(filled_dailysteps, Data = "Imputed Missing Values"),
                             mutate(dailysteps, Data = "Original Values"))
ggplot(combined_dailysteps, aes(x = dailysteps, colour = Data)) +
    geom_histogram(binwidth = 1000, position = "identity", alpha = 0.4, size = 1) +
    labs(x = "Daily Steps", title = "Distribution of Total Daily Steps")

```

### The mean and median of total daily steps using Imputed data set

```{r filled_meandailysteps}
filled_mean_dailysteps <- round(mean(filled_dailysteps$dailysteps), 0)
filled_mean_dailysteps
filled_median_dailysteps <- round(median(filled_dailysteps$dailysteps), 0)
filled_median_dailysteps
change_mean_dailysteps <- round(filled_mean_dailysteps - mean_dailysteps, 0)
change_mean_dailysteps
change_median_dailysteps <- round(filled_median_dailysteps - median_dailysteps, 0)
change_median_dailysteps
```

The mean number of daily steps using imputed data was `r format(filled_mean_dailysteps, scientific = FALSE)`.
This is a change of `r change_mean_dailysteps`.

The median was `r format(filled_median_dailysteps, scientific = FALSE)`.
This is a change of `r change_median_dailysteps`.

## Are there differences in activity patterns between weekdays and weekends?
A column was added indicating whether each day was a weekday or weekend
```{r daytpe}
filled_activity <- mutate(filled_activity, daytype = ifelse(as.POSIXlt(date)$wday == 0 | as.POSIXlt(date)$wday == 6, "Weekend", "Weekday"))
```

The daily activity was calculated grouped by interval and Weekend/Weekday
```{r daytype_dailyactivity}
daytype_dailyactivity <- filled_activity %>%
    group_by(interval, daytype) %>%
    summarise(averagesteps = mean(steps, na.rm = TRUE))
```

### Average number of steps taken vs. 5-minute interval by day type
```{r daytype_lineplot}
ggplot(daytype_dailyactivity, aes(x = interval, y = averagesteps)) +
    geom_line() +
    facet_grid(daytype ~ .) +
    labs(x = "5 Minute Interval", y = "average number of steps", title = "Average number of steps taken vs. 5-Minute interval")

```